---
title: "BC Allied Health Workforce Forecasting Analysis"
subtitle: "Trends, Regional Disparities, and 4-Year Forecasts Across BC Health Authorities"
author: "Keriu Pandya"
date: "2026-03-25"
description: "An analysis of allied health workforce supply across British Columbia, featuring regression-based forecasting, vacancy gap identification, and policy recommendations."
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(scales)
library(knitr)

theme_set(theme_minimal(base_size = 13))

ha_colours <- c(

  "Fraser"            = "#2c7bb6",
  "Vancouver Coastal" = "#d7191c",
  "Island"            = "#fdae61",
  "Interior"          = "#abd9e9",
  "Northern"          = "#018571"
)

prof_colours <- c(
  "Physiotherapist"              = "#1b9e77",
  "Occupational Therapist"       = "#d95f02",
  "Speech-Language Pathologist"  = "#7570b3",
  "Pharmacist"                   = "#e7298a",
  "Medical Lab Technologist"     = "#66a61e",
  "Respiratory Therapist"        = "#e6ab02",
  "Dietitian"                    = "#a6761d",
  "Social Worker"                = "#666666"
)
```

## Introduction

British Columbia faces persistent challenges in health workforce planning. As the province's population ages and grows — particularly in high-growth regions like Fraser Health — demand for allied health services continues to outpace supply. The COVID-19 pandemic further strained an already stretched workforce, accelerating burnout and early retirements while disrupting training pipelines.

This analysis examines **8 allied health professions** across **5 Health Authorities** over **15 years** (2010–2024), with three objectives:

1. Identify growth patterns and regional disparities in workforce supply
2. Produce regression-based forecasts through 2028
3. Assess whether COVID-19 had a statistically significant impact on vacancy rates

::: {.callout-tip}
## Interactive Dashboard
Explore the data yourself with filters and drill-downs in the [Interactive Shiny Dashboard](https://keriu-pandya.shinyapps.io/bc-allied-health-forecast/).
:::

## Data Loading & Quality

```{r load-data}
df <- read_csv("data/bc_allied_health_workforce.csv", show_col_types = FALSE)
```

The raw dataset contains **`r nrow(df)` observations** across `r n_distinct(df$profession)` professions and `r n_distinct(df$health_authority)` Health Authorities.

### Missing Value Assessment

```{r missing-values}
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  mutate(`% Missing` = round(Missing / nrow(df) * 100, 1)) %>%
  filter(Missing > 0)

kable(missing_summary, caption = "Missing values by variable")
```

Approximately 3% of values are missing with no systematic pattern — consistent with data-collection gaps rather than structural bias.

### Imputation

Missing values are imputed using the **group-median** (median within each profession × Health Authority group). This is preferred over mean imputation (sensitive to outliers) and global median (ignores real regional variation).

```{r imputation}
df_clean <- df %>%
  group_by(profession, health_authority) %>%
  mutate(
    headcount       = if_else(is.na(headcount),
                              median(headcount, na.rm = TRUE), headcount),
    supply_per_100k = if_else(is.na(supply_per_100k),
                              median(supply_per_100k, na.rm = TRUE), supply_per_100k),
    vacancy_rate    = if_else(is.na(vacancy_rate),
                              median(vacancy_rate, na.rm = TRUE), vacancy_rate)
  ) %>%
  ungroup()
```

Post-imputation: **`r sum(is.na(df_clean$headcount))` remaining NAs** — validation passed.

## Provincial Workforce Trends

```{r provincial-trends, fig.width=10, fig.height=6}
#| fig-cap: "All professions show positive growth, with Social Workers and Occupational Therapists growing fastest. Note the visible COVID dip in 2020–2021."

provincial_trends <- df_clean %>%
  group_by(year, profession) %>%
  summarise(total_headcount = sum(headcount), .groups = "drop")

ggplot(provincial_trends, aes(x = year, y = total_headcount, colour = profession)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_colour_manual(values = prof_colours) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Total Headcount", colour = "Profession") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2))
```

## Growth Comparison: 2010 vs 2024

```{r growth-comparison, fig.width=9, fig.height=5}
#| fig-cap: "Percentage growth in provincial headcount over the 15-year period."

growth_comparison <- df_clean %>%
  filter(year %in% c(2010, 2024)) %>%
  group_by(year, profession) %>%
  summarise(total = sum(headcount), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = total, names_prefix = "y") %>%
  mutate(
    growth_pct = round((y2024 - y2010) / y2010 * 100, 1),
    profession = fct_reorder(profession, growth_pct)
  )

ggplot(growth_comparison, aes(x = growth_pct, y = profession)) +
  geom_col(fill = "#2c7bb6", width = 0.7) +
  geom_text(aes(label = paste0(growth_pct, "%")), hjust = -0.1, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Growth (%)", y = NULL)
```

## Regional Disparities

### Per-Capita Supply by Health Authority (2024)

```{r supply-by-ha, fig.width=11, fig.height=6}
#| fig-cap: "Northern Health consistently shows the lowest per-capita supply — often 30–50% below urban Health Authorities."

supply_2024 <- df_clean %>% filter(year == 2024)

ggplot(supply_2024, aes(x = profession, y = supply_per_100k, fill = health_authority)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = ha_colours) +
  labs(x = NULL, y = "Professionals per 100k Population", fill = "Health Authority") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 9),
        legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 1))
```

### Vacancy Rate Heatmap (2024)

```{r vacancy-heatmap, fig.width=10, fig.height=6}
#| fig-cap: "Red cells indicate critical shortage areas. Northern Health and specialized professions (SLPs, Dietitians) are most affected."

vacancy_2024 <- df_clean %>% filter(year == 2024)

ggplot(vacancy_2024, aes(x = health_authority, y = profession, fill = vacancy_rate)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = percent(vacancy_rate, accuracy = 0.1)), size = 3.2) +
  scale_fill_gradient(low = "#ffffcc", high = "#d7191c", labels = percent_format()) +
  labs(x = NULL, y = NULL, fill = "Vacancy Rate") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

### Estimated Unfilled Positions

```{r unfilled-positions, fig.width=9, fig.height=5}
#| fig-cap: "Even professions with moderate vacancy rates can have large absolute gaps due to workforce size."

unfilled <- df_clean %>%
  filter(year == 2024) %>%
  mutate(
    estimated_demand   = round(headcount / (1 - vacancy_rate)),
    unfilled_positions = estimated_demand - headcount
  ) %>%
  group_by(profession) %>%
  summarise(total_unfilled = sum(unfilled_positions), .groups = "drop") %>%
  mutate(profession = fct_reorder(profession, total_unfilled))

ggplot(unfilled, aes(x = total_unfilled, y = profession)) +
  geom_col(fill = "#d7191c", width = 0.7) +
  geom_text(aes(label = comma(total_unfilled)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Estimated Unfilled Positions", y = NULL)
```

## Regression Forecasts (2025–2028)

Linear regression models are fitted per profession on provincial headcount totals. **COVID years (2020–2021) are excluded** from model fitting to prevent pandemic-induced dips from distorting the underlying growth trend.

```{r regression-models}
provincial_annual <- df_clean %>%
  group_by(year, profession) %>%
  summarise(total_headcount = sum(headcount), .groups = "drop")

training_data <- provincial_annual %>%
  filter(!(year %in% c(2020, 2021)))

models <- training_data %>%
  group_by(profession) %>%
  nest() %>%
  mutate(
    model  = map(data, ~ lm(total_headcount ~ year, data = .x)),
    glance = map(model, glance),
    tidy   = map(model, tidy)
  )
```

### Model Diagnostics

```{r diagnostics}
diagnostics <- models %>%
  select(profession, glance) %>%
  unnest(glance) %>%
  select(Profession = profession, `R²` = r.squared, `Adj. R²` = adj.r.squared,
         `p-value` = p.value, `Residual SE` = sigma) %>%
  mutate(across(c(`R²`, `Adj. R²`), ~ round(., 4)),
         `p-value` = format.pval(`p-value`, digits = 3),
         `Residual SE` = round(`Residual SE`, 1))

kable(diagnostics, caption = "OLS regression diagnostics per profession")
```

### Forecast Results

```{r forecasts}
forecast_years <- tibble(year = 2025:2028)

forecasts <- models %>%
  select(profession, model) %>%
  mutate(
    predictions = map(model, ~ {
      pred <- predict(.x, newdata = forecast_years, interval = "prediction", level = 0.95)
      bind_cols(forecast_years, as_tibble(pred))
    })
  ) %>%
  select(profession, predictions) %>%
  unnest(predictions) %>%
  rename(forecast = fit, lower_95 = lwr, upper_95 = upr) %>%
  mutate(across(c(forecast, lower_95, upper_95), round))

kable(forecasts, caption = "Projected provincial headcounts with 95% prediction intervals",
      format.args = list(big.mark = ","))
```

```{r forecast-panels, fig.width=12, fig.height=10}
#| fig-cap: "Blue = historical data. Red dashed = forecast with 95% prediction intervals (orange band). Grey zone = COVID years excluded from model."

historical_plot <- provincial_annual %>% mutate(type = "Historical")
forecast_plot <- forecasts %>% rename(total_headcount = forecast) %>% mutate(type = "Forecast")

combined <- bind_rows(
  historical_plot %>% select(year, profession, total_headcount, type),
  forecast_plot   %>% select(year, profession, total_headcount, type)
)

ggplot() +
  geom_line(data = filter(combined, type == "Historical"),
            aes(x = year, y = total_headcount), colour = "#2c7bb6", linewidth = 0.8) +
  geom_point(data = filter(combined, type == "Historical"),
             aes(x = year, y = total_headcount), colour = "#2c7bb6", size = 1.2) +
  geom_ribbon(data = forecast_plot,
              aes(x = year, ymin = lower_95, ymax = upper_95),
              fill = "#fdae61", alpha = 0.35) +
  geom_line(data = filter(combined, type == "Forecast"),
            aes(x = year, y = total_headcount),
            colour = "#d7191c", linewidth = 0.8, linetype = "dashed") +
  geom_point(data = filter(combined, type == "Forecast"),
             aes(x = year, y = total_headcount), colour = "#d7191c", size = 1.2) +
  annotate("rect", xmin = 2019.5, xmax = 2021.5, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.25) +
  facet_wrap(~ profession, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Provincial Headcount") +
  theme(strip.text = element_text(face = "bold", size = 10))
```

## COVID Impact Analysis

Welch's t-test (one-sided, α = 0.05) comparing pre-COVID (2015–2019) vs post-COVID (2022–2024) vacancy rates.

**H₁:** Post-COVID vacancy rates are significantly higher than pre-COVID.

```{r t-tests}
pre_covid  <- df_clean %>% filter(year >= 2015, year <= 2019)
post_covid <- df_clean %>% filter(year >= 2022, year <= 2024)

t_test_results <- tibble(profession = unique(df_clean$profession)) %>%
  mutate(
    test = map(profession, ~ {
      pre  <- pre_covid  %>% filter(profession == .x) %>% pull(vacancy_rate)
      post <- post_covid %>% filter(profession == .x) %>% pull(vacancy_rate)
      t.test(post, pre, alternative = "greater")
    }),
    tidied = map(test, tidy)
  ) %>%
  unnest(tidied) %>%
  select(Profession = profession, `Post-COVID Mean` = estimate1,
         `Pre-COVID Mean` = estimate2, `t-statistic` = statistic,
         `p-value` = p.value) %>%
  mutate(
    Difference = round(`Post-COVID Mean` - `Pre-COVID Mean`, 4),
    Significant = if_else(`p-value` < 0.05, "Yes", "No"),
    across(c(`Post-COVID Mean`, `Pre-COVID Mean`), ~ round(., 4)),
    `p-value` = round(`p-value`, 4)
  )

kable(t_test_results, caption = "Welch's t-test results: post-COVID vs pre-COVID vacancy rates")
```

```{r vacancy-comparison, fig.width=10, fig.height=6}
#| fig-cap: "Several professions show statistically significant increases in vacancy rates following the pandemic."

vacancy_comparison <- bind_rows(
  pre_covid  %>% mutate(period = "Pre-COVID (2015\u20132019)"),
  post_covid %>% mutate(period = "Post-COVID (2022\u20132024)")
) %>%
  group_by(profession, period) %>%
  summarise(mean_vacancy = mean(vacancy_rate), .groups = "drop")

ggplot(vacancy_comparison, aes(x = profession, y = mean_vacancy, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Pre-COVID (2015\u20132019)" = "#abd9e9",
                                "Post-COVID (2022\u20132024)" = "#d7191c")) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Mean Vacancy Rate", fill = NULL) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 9),
        legend.position = "bottom")
```

## Limitations

1. **Synthetic data** — modeled after real patterns but not a substitute for administrative records
2. **Linear assumptions** — constant growth rate may not hold; retirement waves and policy shifts introduce non-linearity
3. **Provincial aggregation** — masks sub-regional variation (e.g., Northern Health stagnation vs Fraser growth)
4. **Supply-side only** — no demand modeling (population health needs, utilization projections)
5. **COVID exclusion trade-off** — prevents distortion but discards information about persistent structural changes

## Recommendations

::: {.callout-important}
### 1. Targeted Recruitment for High-Vacancy Professions
Speech-Language Pathologists and Dietitians need expanded training seats, streamlined international credential recognition, and return-of-service agreements.
:::

::: {.callout-important}
### 2. Northern Health Workforce Strategy
Distributed training models, loan forgiveness, and community-embedded practice are more effective than relocation incentives alone.
:::

::: {.callout-warning}
### 3. Post-COVID Vacancy Monitoring
Establish quarterly vacancy tracking rather than annual surveys to enable earlier intervention.
:::

::: {.callout-note}
### 4. Model Refinement
Extend the framework with retirement age distributions, training program enrollment, and interprovincial migration data.
:::

---

*Built with R (tidyverse, broom, scales) and Quarto. Source code on [GitHub](https://github.com/keriupandya/bc-allied-health-forecast).*
